Forecasting short-term hourly Emergency Departement arrivals

### 2022-01-12 paper/paper.tex
a draft from Bahman to pass to Ivan to go through a final round of reading and improving.

  \journal{Which journal? Journal of Service Research/Health Services Research/EJOR/IJF} % Sets Journal name

@@

  \title{Forecasting short-term hourly Emergency Departement arrivals}
    \author[University1]{Bahmna Rostami Tabar\corref{1}}
   \ead{email1@example.com} 
    \author[School of Mathematics and Statistics, University of Glasgow, UK]{Jethro Browell\corref{2}}
   \ead{jethro.browell@glasgow.ac.uk} 
    \author[Centre for Marketing Analytics and Forecasting, Lancaster University, UK]{Ivan Svetunkov\corref{2}}
   \ead{i.svetunkov@lancaster.ac.uk} 
      \address[University1]{Cardiff business school, 3 Colum Drive, CF10 3EU, Cardiff}
    \address[University2]{adress2}
    \address[University3]{adress3}
      \cortext[1]{Corresponding Author}
    \cortext[2]{Equal contribution}

  \begin{abstract}
  The Objective of this work would be to propose a new methodology to forecast short-term hourly forecasting for urgent and emergency care.
  \end{abstract}

 \end{frontmatter}

@@

Forecasting Emergency Department (ED) arrivals is critical for informing staffing and scheduling decisions to meet the needs of patients. An accurate ED demand forecasts contribute to a better decision making process regarding the resources allocation and staffing. This is one of the best ways to optimize resources utilisation and minimise related costs. State of the art practice to optimise personnel scheduling follows the general approach originally presented by \citet{vile2016time}, which recommends that the following steps be taken to roster employees: (i) forecast demand; (ii) convert demand forecasts into staffing requirements; (iii) schedule shifts optimally; and (iv) assign employees to shifts. An accurate demand forecasting is crucial in ED services to depict various courses of action that can result in massive savings in terms of patient lives. Inability to match the staff with the demand might result in patients overcrowding the system which is a serious problem that causes challenges for the patient flow \citep{derlet2002overcrowding}. Also, it is related with increasing length of stay \citep{muhammet2015forecasting}, low patient satisfaction, unexpected return visits to services, increased health care costs, inaccuracy in electronic medical records and other \citep{rostami2020anticipating}.

An accurate forecasting of arrivals by hour of the day enables planners to match staff to meet anticipated patients, reconfigure units and redeploy staff. This has many advantages for both patients, staff and the quality of provided services. Hourly forecasts are required to inform the short-term operational planning for the current and the upcoming shifts of the day. This involves the short-term decision making related to the execution of the delivery process in ED.
The combination of an hourly forecast demand, current staff being occupied, resource availability and waiting times at ED, provide information on the state of the unscheduled care system across the service. Having this full picture enables the delivery managers to focus on the areas that require intervention to enable the most effective delivery of the service to the patients. However, in comparison with lower frequency time series forecasting such as monthly, quarterly and yearly, hourly forecasts are challenging because the noise caused by random variation may overshadow any pattern in the time series. Hourly time series generally exhibit multiple seasonal cycles of different lengths such as hourly, daily, weekly and yearly. They may also express nonstationarity and their profile may change over time. Therefore, an appropriate forecasting model should take these features into account, to accurately predict hourly demand admissions.

There are few studies that look at forecasting hourly arrivals in ED and other hospital services using historical time series data and/or predictors such as patient characteristics , climate factors , holidays. These studies use multiple approaches including Exponential Smoothing \citep{SvetunkovAdam}, Autoregressive Integrated Moving Average (ARIMA) \citep{hyndman2021forecasting}, Autoregressive Conditional Heteroskedasticity (ARCH) \citep{bollerslev1994arch}, Vector Autoregressive model \citep{lutkepohl2013vector}, TBATS \citep{de2011forecasting} and Artificial Neural Networks \citep{hyndman2021forecasting}. There are few limitations in these studies. They only generate the estimated future arrivals as a point forecast (a single number), which does not acknowledge any uncertainty around the number of arrivals. However, reporting uncertainty is critical because the consequences of imperfect staffing are asymmetric. This asymmetry arises because it is preferable to incur a small opportunity cost associated under utilised staff rather than lower service levels if staff levels are insufficient. This is where producing probabilistic forecasts could be beneficial in staffing. The lack of probabilistic forecast is one of the main bottlenecks in the deployment of generated forecasts in the staffing and planning tools in practice. Probabilistic forecasts are necessary to make decisions that balance the cost associated with under and over staffing. In this paper, in adition to generating point forecast, we also produce and evaluate the forecast distribution of hourly ED arrivals, using several forecasting models, which could be used as a risk management tool for planners and decision makers. Moreover, datasets used in some of these studies are relatively small, short length (e.g.~time period of 1 year), which make it challenging to report the forecast accuracy using robust approaches such as time series cross validation and such results might not be reliable. Additionally, most of forecasting methods used in these publications does not consider the multiple seaonality natutre of hourly ED arrivals. Moreover, all previous publicationscovered in this paper fails reproducibility principles.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  We use the family of Generalised Additive Models to forecast short-term hourly hospital admission using that accounts for i) \ldots, ii) \ldots{} iii) , \ldots{} iv) (Jethro to complete)
\item
  We apply advanced dynamic model based on iETS \citep{Svetunkov2019a} and ETSX models with a modification for multiple frequencies;
\item
  We produce probabilistic forecasts, in addition to the point ones, quantifying the uncertainty in future hospital admission;
\item
  We compare the accuracy of our models with the one from the well established benchmarks, such as Prophet, TBATS, Poisson Regression, and exponential smoothing state space model (ETS);
\item
  We provide data and code enabling reproduction and refinement of the proposed approach and benchmarks. The proposed approach could also be generalized to forecast hourly requirements in other services such as the number of incidents or call volumes in clinical desk services.
\end{enumerate}

The rest of the paper is organised as following: section \ref{lit} provides a brief overview of hourly ED arrival forecasting in healthcare. Section \ref{model} starts with \ldots{}

@@

There are many studies that employ models to forecast the admission and arrivals in Emergency Department. The time granularity considered by these studies spans from hourly to yearly. However, given the focus of the paper, we only discuss the hourly ED forecasting. Table \ref{tab:summarylit} summarize the relevant papers.

@@

\caption{\label{tab:summarylit}Summary of studies in forecasting hourly arrivals in Emergency Department}
\centering
\resizebox{\linewidth}{!}{
\fontsize{11}{13}\selectfont
\begin{tabular}[t]{lr>{\raggedright\arraybackslash}p{6em}>{\raggedright\arraybackslash}p{4em}l>{\raggedright\arraybackslash}p{15em}>{\raggedright\arraybackslash}p{4em}ll}
\toprule
Author & Year & Variable & Horizon & Length & Method & Metric & Probabilistic & Multiple seasonality\\
\midrule
Choudhury and Urena & 2020 & ED arrivals & 1h to 24h & 4 & ARIMA; Holt-winters; TBATS; ANN & RMSE, ME & No & Yes\\
Asheim et al. & 2019 & ED arrivals & 3h & 5 & Poisson regression & MAPE & No & No\\
Hertzum & 2017 & ED arrivals & 1,2,4,8,24 hours & 3 & linear regression; SARIMA; Naïve & MAE, MAPE, 
MASE & No & No\\
Kim et al. & 2014 & Hospital admission & 4h, 24h & 3 & Linear regression; Exponential smoothing; ARIMA; GARCH; VAR & MAPE & No & No\\
Cote et al. & 2013 & ED arrivals & 24h & 2 & Fourier regression & R\textasciicircum{}2,  Standard Error & No & No\\
Chase et al. & 2012 & ED CUR & 30m
1h, 2h, 4h, 8h, 12h & 1 & Binary regression & NA & No & No\\
Schweigler et al. & 2009 & Bed occupancy & 4h, 12h & 4 & Hourly historical average; SARIMA; Sinusoidal model with autocorrelated error & RMSE & No & No\\
Jones et al. & 2009 & ED census & 24h & 2 & VAR;  Holt winters & MAE & No & No\\
McCarthy & 2008 & ED arrivals & 24h & 2 years & Poisson log-linear regression model & 95\% CI & No & Yes\\
Channouf et al. & 2007 & Ambulance admission & 1h, 3h, 6h, 12h, 13h, 14h, 17h, 23h, 24h & 2 & Regression & RMSE & No & No\\
Morzuch and Allen & 2006 & ED arrivals & 168h & 3 & Regression; ARIMA; Exponential smoothing & RMSE & No & No\\
\bottomrule
\end{tabular}}
\end{table}
\end{landscape}

Linear regression, ARIMA, and Naïve were used by \citet{hertzum2017forecasting} to investigate whether accurate hourly accident and emergency department patient arrivals and occupancy forecasts can be generated using calendar variables. Naïve method was there for the purpose of comparison. \citet{hertzum2017forecasting} study shows that patient arrivals variation is larger across the hours of the day than across the days of the week and the months of the year. In terms of hour of the day, patient arrivals peaked around noon. For days of the week, Monday is the busiest day while weekends are the quietest ones. July-August are the months with the highest numbers of patient arrivals, while January and February are the months with the lowest numbers. According to \citet{hertzum2017forecasting}, the regression and ARIMA models performed similarly in modelling patient arrivals, while ARIMA outperformed regression models in modelling accident and emergency department occupancy. However, after all, the models of occupancy were less accurate than those of arrivals. \citet{hertzum2017forecasting} mentioned that ARIMA was among the most accurate models in terms of forecasting accident and emergency department visits.

\citet{choudhury2020forecasting} used ARIMA, Holt-Winters, TBATS, and neutral network methods to forecast hourly accident and emergency department arrivals. ARIMA model was selected as the best fit model. Authors claimed that ARIMA has provided high and acceptable hourly ED forecasting accuracy, even outperforming TBATS.

\citet{morzuch2006forecasting} used the Unobserved Components Model (UCM), in which each component of the time series is separately modeled as stochastic. Double-seasonal exponential smoothing and standard Holt-Winters were used to forecast ED arrival for an horizon of 168 hours. The hourly data collected from an ED in Pennsylvania showed no trend, and two seasonal cycles: a within-day and a within-week seasonal cycles. The double seasonal model recorded lower RMSEs for all the 168-hour horizons, which was expected due to the strong hourly seasonality of the time series.

\citet{mccarthy2008challenge} employed a Poisson log-linear regression model, including independent variables such as temporal factors (e.g., hour-of-day, day- of-week, type-of-day, season, and calendar year), patient characteristics (i.e., age, gender, insurance status, triage level, mode of arrival, and ambulance diversion status) and climatic factors (i.e., temperature and precipitation) to forecast patient demand for ED services. Hourly data of ED arrivals in the 1-year study period was deployed to forecast from 1 hour to 24 hours into the future. The authors presented the prediction intervals with \(50\%\) and \(90\%\) confidence levels for the number of hourly arrivals under the Poisson assumption. They showed that the most important predictor was hour of the day and AR(1).

\citet{kim2014predicting} compared different univariate and multivariate time series forecasting techniques to forecast patient volume for a Hospital Medicine programme. The study evaluated linear regression, exponential smoothing, ARIMA, SARIMA, GARCH (generalized autoregressive conditional heteroskedasticity method) and vector autoregressive (VAR) models to forecast for 4 hours and 24 hours ahead. They used MAPE to report the forecast accuracy. ARIMA model outperformed all the other models.

@@

According to the studies mentioned above, it can be said that they have shown complications in forecasting hourly patient accident and emergency department visits and the application of forecasting hourly patients visits is not well established. Some of the studies claimed that the accuracy of forecasting models on hourly accident and emergency department data forecasting model is low compared to the higher forecasting intervals, like daily \citep{boyle2012predicting, hertzum2017forecasting}. However, some studies mentioned that the accuracy of ED hourly forecast is at the acceptable level \citep{choudhury2020forecasting, mccarthy2008challenge, schweigler2009forecasting}.

@@

\begin{enumerate}
\def\labelenumi{(\roman{enumi})}
\tightlist
\item
  Current approaches to forecast hourly ED arrivals do not fully consider the feature of data such as multiple seasonal cycles and changing profile over time;
\item
  Almost all research studies produce point forecasts and consequently report only point forecast accuracy. There is a lack of studies presenting probabilistic forecasts of hourly ED arrivals that better represent uncertainty of future admissions, providing a holistic picture of future demand for a planner;
\item
  most studies are not reproducible, as it is almost impossible to reapply the approaches without the helpf of the authors of those papers;
\item
  studies are limited in terms of the length of historical data used for training purposes and forecast performance evaluation and
\item
  some studies in this area lack a rigorous experimental design, i.e.~there is no benchmark method or forecast accuracy is not reported.
\end{enumerate}

@@

Data used in this study comprises counts of patients' arrival times at one of the largest ED units in the UK between April 2014 and February 2019, extracted from the ED administrative database of the hospital. We aggregated the patients' arrival times to obtain hourly arrivals, which is used for empirical evaluation in this study.

Figure \ref{fig:hourly-plot-ridge} illustrates the distribution of arrivals for each hour of the day and the day of the week. Although the data is noisy, it reveals some systematic patterns.

@@

It is clear that the number of arrivals has a sub-daily structure, similar to the one summarised by \citet{hertzum2017forecasting}. The ED arrivals decrease between mid-night and early morning and then increase until the evening, decreasing after that again. It is also clear that ED service gets systematically more visits on Mondays between 8 a.m. and 5 p.m. Moreover, The number of arrivals around mid-night is slightly higher for Saturday and Sunday.

Figure \ref{fig:hourly-plot-ridge} also highlights that there are many ouliers for almost every hour of the day that may affect the accuracy of forecasting methods. Some of these outliers might be related to holidays and special events. It is also clear that arrivals are less volatile between mid-night and early morning.

@@

\textbf{IS: We need to plot day of week and week of year seasonalplots here and discuss. This will show multiple seasonality nature of the data.}

@@

@@

Based on this simple analysis, we should consider model that could take the following into account:

\begin{itemize}
\tightlist
\item
  Hour of day, day of week, and week of year seasonal patterns,
\item
  Calendar events, such as holidays,
\item
  Lags of calendar events to accommodate the potential changes in demand the next day after a holiday,
\item
  Potentially changing level of data,
\item
  Other events, such as football matches in the area.
\end{itemize}

@@

We start with one of the simplest forecasting approaches used in practice - the one that assumes that in the next few hours, everything will be the same as in the similar hours of a similar day in the past. This is called ``Naïve''. In our case, given that we need a distribution of values, we will use a modified approach, used in climatology, where the empirical distribution of demand is used to forecast the future demand distribution. \textbf{We need references here\ldots{}}

\hypertarget{multiple-regression}{%
\subsection{Multiple Regression}\label{multiple-regression}}

Regression is one of the most popular forecasting methods that uses explanatory variables to predict a variable of interest (in our case, the ED arrivals). The classical linear regression model is formulated as:

@@

where \(\mathbf{x}_t\) is the vector of explanatory variables, \(\boldsymbol{\beta}\) is the vector of parameters, \(\epsilon_t\) is the error term, which is typically assumed to follow Normal distribution with zero mean and a fixed variance, and \(t\) is the time index. However, in the context of healthcare and ED arrivals, the assumption of Normality is unrealistic, because the number of admitted patients is integer and non-negative. So the linear regression model should be substituted by some other model. One of the models that is frequently used in practice is the Poisson regression \citep[see for example,][]{mccarthy2008challenge}, which can be summarised as:

\begin{equation}
  \begin{aligned}
    & {y}_t \sim \mathrm{Poisson} \left( \lambda_t \right) \\
    & \log \lambda_t = \mathbf{x}_t' \boldsymbol{\beta}
  \end{aligned}
\label{eq:PoissonRegression}
\end{equation}

The logarithm in \eqref{eq:PoissonRegression} is needed in order to make sure that the parameter of Poisson distribution is always positive. This model can be estimated via maximisation of the likelihood function based on Poisson mass function {[}reference needed{]}. When it comes to selecting explanatory variables for the model, there is no single correct answer, and the decision needs to be done for each specific case. In our experiment, we will only include dummy variables, capturing a variety of calendar events:

@@

As our data are counts, the natural starting point is the Poisson distribution with an additive model for \(\log \lambda_t\) of the form:

@@

Truncated Normal\ldots{}

with

>>>>

  \journal{Annals of Emergency Medicine} % Sets Journal name

@@

  \title{Forecasting short-term hourly Emergency Departement arrivals}
    \author[Cardiff Business School]{Bahmna Rostami Tabar\corref{12}}
   \ead{rostami-tabarb@cardiff.ac.uk} 
    \author[University of Glasgow]{Jethro Browell\corref{1}}
   \ead{jethro.browell@glasgow.ac.uk} 
    \author[Lancaster University]{Ivan Svetunkov\corref{1}}
   \ead{i.svetunkov@lancaster.ac.uk} 
      \address[Cardiff Business School]{Cardiff Business School, Cardiff University, UK}
    \address[University of Glasgow]{School of Mathematics and Statistics, University of Glasgow, UK}
    \address[Lancaster University]{Centre for Marketing Analytics and Forecasting, Lancaster University, UK}
      \cortext[12]{Corresponding Author}
    \cortext[1]{Equal contribution}

  \begin{abstract}
  An accurate forecast of Emergency Department (ED) arrivals by hour of the day is cirtical to meet patients's demand It will enable planners to match ED staff to number of arrivals, redeploy staff and reconfigure units, if required. This can have many advantages for healthcare staff and the quality of care delivered to patients. In this study, we develop an innovative model based on Generalised Additive Models and an advanced dynamic model based on exponential smoothing, to generate hourly probabilistic forecast of ED arrivals. We compare the forecast accuracy of these models against appropriate benchmarks including TBATS, Poisson Regresion, Prophet and a simple empirical distribution. We use Root Mean Squared Error (RMSE) to examin the point forecast accuracy and the forecast distribution accuracy is assessed using Quantile Bias, PinBall Score and Pinball Skill Score. Our results indicate that the proposed models outperform their benchmarks for both point and probabilistic forecasts. Our developed models can also be generalised to forecast hourly arrivals in other services such as hospitals, ambulance or clinical desk services.
  \end{abstract}
   \begin{keyword} Emergency Department, Arrivals, Poisson Regression, Probabilistic Forecasting, Generalised Additive Models, Intermittent Exponential Smoothing\end{keyword}
 \end{frontmatter}

@@

Forecasting Emergency Department (ED) arrivals is critical for informing staffing and scheduling decisions to meet the needs of patients. An accurate ED demand forecasts contribute to a better decision making process regarding the resources allocation and staffing. This is one of the best ways to optimize resources utilisation and minimise related costs.
An accurate forecast of patient arrivals is crucial in ED services to depict various courses of action that can result in massive savings in terms of patient lives. Inability to match the staff with the demand might result in patients overcrowding the system which is a serious problem that causes challenges for the patient flow \citep{derlet2002overcrowding}. Also, it is related with increasing length of stay \citep{muhammet2015forecasting}, low patient satisfaction, unexpected return visits to services, increased health care costs, inaccuracy in electronic medical records and other \citep{rostami2020anticipating}.

An accurate forecasting of arrivals by hour of the day enables planners to match staff to meet anticipated patients, reconfigure units and redeploy staff. This has many advantages for both patients, staff and the quality of provided services. Hourly forecasts are required to inform the short-term operational planning for the current and the upcoming shifts of the day. This involves the short-term decision making related to the execution of the delivery process in ED.
The combination of an hourly arrival forecast, current staff being occupied, resource availability and waiting times at ED, provide information on the state of the unscheduled care system across the service. Having this full picture enables the delivery managers to focus on the areas that require intervention to enable the most effective delivery of the service to the patients. However, in comparison with lower frequency time series forecasting such as monthly, quarterly and yearly, hourly forecasts are challenging because the noise caused by random variation may overshadow any pattern in the time series. Hourly time series generally exhibit multiple seasonal cycles of different lengths such as hourly, daily, weekly and yearly. They may also express nonstationarity and their profile may change over time. Therefore, an appropriate forecasting model should take these features into account, to accurately predict hourly demand admissions.

There are few studies that look at forecasting hourly arrivals in ED and other hospital services using historical time series data and/or predictors such as patient characteristics, weather, holidays and public events. These studies use multiple approaches including Exponential Smoothing \citep{SvetunkovAdam}, Autoregressive Integrated Moving Average (ARIMA) \citep{hyndman2021forecasting}, Autoregressive Conditional Heteroskedasticity (ARCH) \citep{bollerslev1994arch}, Vector Autoregressive model \citep{lutkepohl2013vector}, TBATS \citep{de2011forecasting} and Artificial Neural Networks \citep{hyndman2021forecasting}. However, majority of these studies are limited to only predict future arrivals as a point forecast (a single number), which does not quantify any uncertainty associated with the number of future arrivals. There are few studies that report uncertainty by presenting prediction intervals, no study generates and evaluates the entire forecast distribution of arrivals. Reporting the uncertainty via the forecast distribution is potentially valuable in this setting because the consequences of imperfect staffing are asymmetric. This asymmetry arises because it is preferable to incur a small opportunity cost associated under utilised staff rather than compromise service levels, if staff levels are insufficient. Probabilistic forecasts inform decision-makers about exposure to these risks and potentially enable those risks to be managed more efficiently. Furthermore, if the impact of under and over staffing can be quantified, probabilistic forecasts allow `optimal' decisions that balance the cost associated with under and over staffing to be calculated. Therefore, in this paper, in addition to generating point forecasts, we also produce and evaluate density forecasts of hourly ED arrivals, comparing several methods for this task. Another drawback of existing studies is that the datasets used are relatively small (e.g.~time period of 1-2 year), which make it challenging to report the forecast accuracy using robust approaches such as time series cross validation and such results might not be generalisable. Additionally, most of forecasting methods used in these publications does not consider the full extend of multiple seasonality exhibited by hourly ED arrivals. Moreover, all previous publications referenced in this paper are not fully reproducible as underlying data and code are not available.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  We produce probabilistic forecasts, in addition to the point estimation, which quantify uncertainties in future hospital admission, and compare different forecasting methods using a suite of established evaluation metrics;
\item
  We develop an advanced dynamic model to forecasts ED arrivals based on iETS \citep{Svetunkov2019a} and ETSX models with a modification for multiple frequencies which produced highly-accurate point forecasts;
\item
  We develop a novel model to produce probabilistic forecast of ED arrivals based on Generalised Additive Models for Location Scale and Shape, which accounts for i) the bounded and non-Gaussian distribution of arrivals, ii) multiple seasonalities, weather and holiday effects, and iii) variation in forecast uncertainty;
\item
  We benchmark the accuracy of our model against appropriate models used when multiple seasonality is present, i.e.~Prophet, TBATS, Poisson Regression, Exponential Smoothing State Space model (ETS) and a simple empirical distribution of the arrivals;
\item
  We provide data and code enabling reproduction and refinement of the proposed approach and benchmarks. The proposed approach could also be generalised to forecast hourly requirements in other services such as the number of incidents or call volumes in clinical desk services.
\end{enumerate}

The rest of the paper is organised as following: In section \ref{lit}, we provide a brief overview of hourly ED arrival forecasting; In Section \ref{data}, we present the hourly time series of an ED arrival and use various plots to highlight important patterns. In section \ref{model}, we describe the modelling approach and benchmark methods. We then discuss the performance evaluation metrics in section \ref{accuracy}; in section \ref{result}, we present and discuss our results. Finally, we summarise our findings and present ideas for future research in section \ref{conclusion}.

@@

There are some studies that employ models to forecast admissions and arrivals in Emergency Department. The time granularity considered by these studies spans from hourly to yearly. However, given the focus of the paper, we only discuss the hourly ED forecasting.

Linear regression, ARIMA, and Naïve were used by \citet{hertzum2017forecasting} to investigate whether accurate hourly accident and emergency department patient arrivals and occupancy forecasts can be generated using calendar variables. \citet{hertzum2017forecasting} study found that patient arrival variation is larger across the hours of the day than across the days of the week and the months of the year. In terms of hour of the day, patient arrivals peaked around noon. For days of the week, Monday is the busiest day while weekends are the quietest ones. July-August are the months with the highest numbers of patient arrivals, while January and February are the months with the lowest numbers. They indicate that regression and ARIMA models performed similarly in modelling patient arrivals, while ARIMA outperformed regression models in modelling accident and emergency department occupancy.

\citet{choudhury2020forecasting} used ARIMA, Holt-Winters, TBATS, and neutral network methods to forecast hourly accident and emergency department arrivals. ARIMA model was selected as the best fit model. Authors claimed that ARIMA has provided high and acceptable hourly ED forecasting accuracy, even outperforming TBATS. \citet{Cheng2021} develop an ARIMA model for ED occupancy with a seasonal component and exogenous variables, which is found to outperform a rolling-average benchmark. They also produce prediction intervals, a form of probabilistic forecast, which are found to be well calibrated, a necessary property for such forecasts.

\citet{morzuch2006forecasting} used the Unobserved Components Model (UCM), in which each component of the time series is separately modelled as stochastic. Double-seasonal exponential smoothing and standard Holt-Winters were used to forecast ED arrival for an horizon of 168 hours. The hourly data collected from an ED in Pennsylvania showed no trend, and two seasonal cycles: a within-day and a within-week seasonal cycles. The double seasonal model recorded lower RMSEs for all the 168-hour horizons, which was expected due to the strong hourly seasonality of the time series.

\citet{mccarthy2008challenge} employed a Poisson log-linear regression model, including independent variables such as temporal factors (e.g., hour-of-day, day- of-week, type-of-day, season, and calendar year), patient characteristics (i.e., age, gender, insurance status, triage level, mode of arrival, and ambulance diversion status) and climatic factors (i.e., temperature and precipitation) to model patient demand for ED services. The authors produced probabilistic predictions in the form of \(50\%\) and \(90\%\) prediction intervals for the number of hourly arrivals. Hourly data of ED arrivals in the 1-year study period was modelled and analysed, and it is suggested that the model could be used for forecasting, however, model evaluation was performed in-sample on only one year of data, so unclear how this approach would perform in a forecasting setting or compare to simpler approaches. However, the length of time series is very short (only one year), which does not allow for a rigorous out-of-sample evaluation.

\citet{kim2014predicting} compared different univariate and multivariate time series forecasting techniques to forecast patient volume for a Hospital Medicine programme. The study evaluated linear regression, exponential smoothing, ARIMA, SARIMA, Generalized Autoregressive Conditional Heteroskedasticity (GARCH) and Vector Autoregressive (VAR) models to forecast for 4 hours and 24 hours ahead. They used Mean Absolute Percentage Error (MAPE) to report the forecast accuracy. ARIMA model outperformed all the other models.

Table \ref{tab:summarylit} summarize the relevant papers.

@@

\caption{\label{tab:summarylit}Summary of studies in forecasting hourly arrivals in Emergency Department}
\centering
\resizebox{\linewidth}{!}{
\fontsize{11}{13}\selectfont
\begin{tabular}[t]{lr>{\raggedright\arraybackslash}p{6em}>{\raggedright\arraybackslash}p{4em}l>{\raggedright\arraybackslash}p{10em}>{\raggedright\arraybackslash}p{8em}ll}
\toprule
Author & Year & Variable & Horizon & Length & Method & Metric & Probabilistic & Seasonality\\
\midrule
Cheng et al. & 2021 & ED visits & 1h to 4 h & 1 year & SARIMAX, Holt-Winters, VAR, ARIMA & MSE, MAE, MAPE, Prediction interval coverage & NO & Single\\
Choudhury and Urena & 2020 & ED arrivals & 1h to 24h & 4 years & ARIMA; Holt-winters; TBATS; ANN & RMSE, ME & No & Multiple\\
Asheim et al. & 2019 & ED arrivals & 3h & 5 years & Poisson regression & MAPE & No & Single\\
Hertzum & 2017 & ED arrivals & 1,2,4,8,24 h & 3 years & linear regression; SARIMA; Naïve & MAE, MAPE, 
MASE & No & Single\\
Kim et al. & 2014 & Hospital admission & 4h, 24h & 3 years & Linear regression; Exponential smoothing; ARIMA; GARCH; VAR & MAPE & No & Single\\
Cote et al. & 2013 & ED arrivals & 24h & 2 years & Fourier regression & R\textasciicircum{}2,  Standard Error & No & Single\\
Chase et al. & 2012 & ED CUR & 30m
1h, 2h, 4h, 8h, 12h & 1 year & Binary regression & NA & No & Single\\
Schweigler et al. & 2009 & Bed occupancy & 4h, 12h & 4 years & Hourly historical average; SARIMA; Sinusoidal model with autocorrelated error & RMSE & No & Single\\
Jones et al. & 2009 & ED census & 24h & 2 years & VAR;  Holt winters & MAE & No & Single\\
McCarthy & 2008 & ED arrivals & n/a & 1 year & Poisson log-linear regression model & Prediction interval coverage & Yes & Multiple\\
Channouf et al. & 2007 & Ambulance admission & 1h, 3h, 6h, 12h, 13h, 14h, 17h, 23h, 24h & 2 years & Regression & RMSE & No & Single\\
Morzuch and Allen & 2006 & ED arrivals & 168h & 3 years & Double Exponential Smoothing; Additive Holt Winter & RMSE & No & Multiple\\
\bottomrule
\end{tabular}}
\end{table}
\end{landscape}

@@

\citet{cheng2021forecasting} used one year of ED visits time series to evaluate the performance Rolling Average, SARIMAX, ARIMA, VAR and Holt-Winter to forecast ED occupancy up to 4-hours ahead. The forecast accuracy is evaluated using Mean Squared Error (MSE), Mean Absolute Error (MAE) and MAPE for point forecast and coverage for prediction intervals of \(80\%\) and \(95\%\). They show that SARIMAX provides more accurately forecast of hourly ED occupancy.

According to the studies mentioned above, it can be said that they have shown complications in forecasting hourly patient accident and emergency department visits and the application of forecasting hourly patients visits is not well established. Some of the studies claimed that the accuracy of forecasting models on hourly accident and emergency department data forecasting model is low compared to the higher forecasting intervals, like daily \citep{boyle2012predicting, hertzum2017forecasting}. While, others mentioned that the accuracy of ED hourly forecast is at the acceptable level \citep{choudhury2020forecasting, mccarthy2008challenge, schweigler2009forecasting}.

@@

\begin{enumerate}
\def\labelenumi{(\roman{enumi})}
\tightlist
\item
  Current approaches to forecast hourly ED arrivals do not fully consider the feature of data such as multiple seasonal cycles and changing profile over time;
\item
  Almost all research studies produce point forecasts and at best report prediction intervals. There is a lack of studies presenting the entire forecast distribution of hourly ED arrivals that better represent uncertainty of future arrivals, providing a holistic picture of future demand for a planner;
\item
  most studies are not reproducible, as it is almost impossible to reapply the approaches without the help of the authors of those papers;
\item
  studies are limited in terms of the length of historical data used for training purposes and forecast performance evaluation and
\item
  some studies in this area lack a rigorous experimental design, i.e.~there is no benchmark method or forecast accuracy is not reported.
\end{enumerate}

@@

Data used in this study comprises counts of patients' arrival times at one of the largest ED units in the UK between April 2014 and February 2019, extracted from the ED administrative database of the hospital. We aggregated the patients' arrival times to obtain hourly arrivals, which are used in this study. Figure \ref{fig:hourly-plot-ridge} illustrates the distribution of arrivals for each hour of the day and the day of the week. Although the data is noisy, it reveals some systematic patterns.

@@

It is clear that the number of arrivals has a sub-daily structure, similar to the one summarised by \citet{hertzum2017forecasting}. The ED arrivals decrease between mid-night and early morning and then increase until the evening, decreasing after that again. It is also clear that ED service gets systematically more visits on Mondays between 8 a.m. and 5 p.m. Moreover, the number of arrivals around mid-night is slightly higher for Saturday and Sunday.

Figure \ref{fig:hourly-plot-ridge} also highlights that there is significant skewness for almost every hour of the day that varies with time-of-day, and which should be accounted for in forecasting methods. Some skewness might be related to holidays and special events. It is also clear that arrivals are less volatile between mid-night and early morning.

@@

Figure \ref{fig:seasonplot-weekofyear} illustrates the daily seasonal plot, with x-axis representing the day-of-week and y-axis the ED arrivals. There is a line for each week (i.e.~7 days) from Monday to Saturday, and lines are color-coded by week of year to show the weekly cycle. It is clear that there is a large jump in patient arrivals on Monday followed by Saturday each week. This shows that there are significantly more arrivals on Mondays and Saturdays comparign to the rest of the week. This might be due to the closure of General Practitioners outpatient clinics over the weekend.

@@

Figure \ref{fig:hourly-plot-ridge} highlights the week of year seasonality in the ED arrivals. We observe that arrivals are significantly lower from week 29 to 35, which corresponds to the summer period. Moreover, the number of arrivals are lower at beginning and the end of the year. The highest arrivals are in week 39-42.

@@

Figure \ref{fig:hourly-plot-ridge} presents the time plot of daily arrivals. Each point represents one day, and points are shape-coded by day-of-week to show the weekly cycle. The figure shows more arrivals on Mondays compared to other weekdays. Moreover, we can observe a long-term trend line as well as the significant effect of some holidays. We can see that arrivals near Christmas and New Year's day are significantly lower than other days of the year. Moreover, Figure \ref{fig:hourly-plot-ridge} allows us to identify the impact of special event on arrivals. For instance, we see that the number of arrivals are significamtly low for 01-03 of March 2018. These days correspond to the Storm Emma with heavy snowfall that resulted in travel disruption, mass power outages and schools closed in the UK \citep{stormemma2018}.

Based on ED time series analysis illustrated above and literature review, we should consider models that can take the following into account:

\begin{itemize}
\item
  Daily, weekly, and annual seasonalities,
\item
  Long-term trend,
\item
  Calendar events, such as holidays,
\item
  Lags of calendar events to accommodate the potential changes in demand the next day after a holiday,
\item
  Other events, such as sporting fixtures,
\item
  Weather effects, such as temperature and precipitation.
\end{itemize}

@@

We start with one of the simplest forecasting approaches used in practice - the one that assumes that in the next few hours, everything will be the same as in the similar hours of a similar day in the past. This is called ``Naïve''. In our case, given that we need a distribution of values, we will use a modified approach, used in climatology, where the empirical distribution of the hourly arrival time series is used to forecast the future arrival distribution \citep{la2021new}.

\hypertarget{poisson-regression}{%
\subsection{Poisson Regression}\label{poisson-regression}}

Regression is one of the most popular forecasting methods that uses explanatory variables to predict a variable of interest (in our case, the ED arrivals). The classical linear regression model is formulated as

@@

where \(\mathbf{x}_t\) is the vector of explanatory variables, \(\boldsymbol{\beta}\) is the vector of parameters, \(\epsilon_t\) is the error term, which is typically assumed to follow Normal distribution with zero mean and a fixed variance, and \(t\) is the time index. However, in the context of healthcare and ED arrivals, the assumption of Normality is unrealistic, because the number of admitted patients is integer and non-negative. So the linear regression model should be substituted by some other model. One of the models that is frequently used in practice is the Poisson regression \citep[see for example,][]{mccarthy2008challenge}, which can be summarised as

\begin{equation}
  \begin{aligned}
    & {y}_t \sim \mathrm{Poisson} \left( \lambda_t \right) \\
    & \log \lambda_t = \mathbf{x}_t' \boldsymbol{\beta}
  \end{aligned} \quad .
\label{eq:PoissonRegression}
\end{equation}

The logarithm in \eqref{eq:PoissonRegression} is needed in order to make sure that the parameter of Poisson distribution is always positive. This model can be estimated via maximisation of the likelihood function based on Poisson mass function . When it comes to selecting explanatory variables for the model, there is no single correct answer, and the decision needs to be done for each specific case. In our experiment, we will only include dummy variables, capturing a variety of calendar events:

@@

As our data are counts, the natural starting point is the Poisson distribution with an additive model for \(\log \lambda_t\) of the form

@@

However, experiments on the training data reveal that calibration of forecasts based on the Poisson distribution is poor, suggesting that the shape of the distribution is unsuitable for the present application. In particular, we observe that forecast uncertainty appears to vary depending on the time of day and possibly other explanatory variables. Therefore, we consider more flexible, two parameter distributions so that we are able to specify additive models for both location and scale parameters, specifically the truncated Normal distribution, with truncation at 0. The resulting density forecasts are given by
